---
title: "Chun Shing Leung TidyVerse Assigment"
author: "CHUN SHING LEUNG"
date: "2023-11-11"
output: html_document
---

Your task here is to Create an Example.  Using one or more TidyVerse packages, and any dataset from fivethirtyeight.com or Kaggle, create a programming sample “vignette” that demonstrates how to use one or more of the capabilities of the selected TidyVerse package with your selected dataset.

## Library & Data

The dataset is selected from Kaggle:

https://www.kaggle.com/datasets/jessemostipak/hotel-booking-demand/code

https://raw.githubusercontent.com/tonyCUNY/FALL2023TIDYVERSE/main/hotel_bookings.csv

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
hotel <- read_csv("https://raw.githubusercontent.com/tonyCUNY/FALL2023TIDYVERSE/main/hotel_bookings.csv")

```

## TidyVerse Package

Tidyverse consists of the following pacakges. 

ggplot2 - A system for declaratively creating graphics, based on The Grammar of Graphics.
dplyr - It provides a grammar of data manipulation, providing a consistent set of verbs that solve the most common data manipulation challenges
tidyr -  It provides a set of functions that help you get to tidy data.
readr - It provides a fast and friendly way to read rectangular data (like csv, tsv, and fwf)
purrr - It enhances R’s functional programming (FP) toolkit by providing a complete and consistent set of tools for working with functions and vectors.
tibble - A modern re-imagining of the data frame, keeping what time has proven to be effective, and throwing out what it has not.
stringr - It provides a cohesive set of functions designed to make working with strings as easy as possible.
forcats - It provides a suite of useful tools that solve common problems with factors. 

Dplyr and ggplot will be demonstrated

## Dplyr

```{r  warning=FALSE, message=FALSE}

# Use |>  to emphasis a sequence of actions, rather than the object that the actions are being performed on.
# drop_na() - drops rows contains missing value. 
# distinct() - Keep distinct/unique rows
# select() - Keep or drop columns using their names and types
# mutate() - Create, modify, and delete columns
# filter() - Keep rows that match a condition
# relocate() - Change column order
# group_by() - Group by one or more variables
# summarise() - Summarise each group down to one row
# glimpse - Get a glimpse of your data

hotel2 <- hotel |> 
              drop_na() |> 
              distinct() |> 
              select(-c(agent, company, adr)) |> 
              mutate(arrival_date_month = match(tolower(arrival_date_month), tolower(month.name))) |> 
              mutate(arrival_date = as.Date(paste(arrival_date_year, arrival_date_month, arrival_date_day_of_month, sep = "-"))) |> 
              select(-arrival_date_week_number) |> 
              filter(adults != 0) |> 
              mutate(kids = children + babies) |>
              select(-c(children, babies)) |> 
              relocate(kids, .after = adults) |> 
              mutate(total_nights = stays_in_weekend_nights + stays_in_week_nights) |>
              select(-c(stays_in_weekend_nights, stays_in_week_nights)) |> 
              relocate(total_nights, .after = arrival_date_day_of_month)

hotel3 <- hotel2 |>
              group_by(country) |> 
              summarise(count = n())

glimpse(hotel2)

```

## GGplot - Barplot

```{r warning=FALSE, message=FALSE}
# aes() - It maps variables from the dataset to visual properties (aesthetics) of geometric objects (geoms) in the plot
# geom_bar() - It makes the height of the bar proportional to the number of cases in each group (or if the weight aesthetic is supplied, the sum of the weights).
# coord_flip() - flip the coordinate
# labs() - Modify axis, legend, and plot labels

# Top 10 Countries of Origin for Hotel Visitors

hotel3 |> 
  top_n(10) |> 
  ggplot(aes(x = reorder(country, count), y = count, fill = country)) +
  geom_bar(stat = "identity") + 
  coord_flip() +
  labs(title = "Top 10 Countries of Origin for Hotel Visitors",
       x = "Country",
       y = "Number of Guests")



```
