---
title: "tidyverse"
author: "Michael Robinson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction 

In this assignment, I will explore a synthetic healthcare dataset using the TidyVerse packages. The dataset, designed to mimic real-world healthcare data, is comprehensive, containing 10,000 rows and 15 columns. It includes a range of information typical in healthcare settings, such as patient names, ages, genders, medical conditions, admission and discharge dates, and more.
thtih data was obtain from https://www.kaggle.com/datasets/prasad22/healthcare-dataset


# Load the Dataset and Libraries needed

 
```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(DT)

patient_data <- read_csv("https://raw.githubusercontent.com/MRobinson112/tidy_verse/main/healthcare_dataset.csv")


datatable(patient_data, options = list(scrollX = TRUE))


```

# Filtering

Filtering is important in data analysis, it is used to narrow dataset to the relevant data. This increases the accuracy and relevance of the analysis by removing unnecessary data.

```{r}

# filter all patients over 50
patients_over_50 <- patient_data %>%
  filter(Age > 50
         
# filter all patients over 50 that are male 
male_patients_over_50 <- patient_data %>%
  filter(Gender == "Male", Age > 50)
```
# Arranging

Arranging data is used in data analysis to aid in understanding trends and patterns.

```{r}

# sort patientes by date of admission 

patients_sorted_by_admission <- patient_data %>%
  arrange(`Date of Admission`)

head(patients_sorted_by_admission)
```
# Summerizing

Summarizing is important in data analysis because it helps to manage large datasets, find patterns, compare different data, and make decisions.

```{r}

average_billing_by_condition <- patient_data %>%
  group_by(`Medical Condition`) %>%
  summarize(Average_Billing = mean(`Billing Amount`))

average_billing_by_condition
```

# Visualization 

visualization transforms data into a more accessible and interpretable form, enhancing understanding, communication, and decision-making.

```{r}

# Count of male patients over 50
count_male_over_50 <- nrow(male_patients_over_50)

# Count of all patients over 50
count_patients_over_50 <- nrow(patients_over_50)

# create a new data frame of summarized information.

plot_data <- data.frame(
  Category = c("Male Patients Over 50", "All Patients Over 50"),
  Count = c(count_male_over_50, count_patients_over_50)
)

ggplot(plot_data, aes(x = Category, y = Count, fill = Category)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Comparison of Male Patients Over 50 vs. All Patients Over 50",
       x = "",
       y = "Count")

```
