---
title: "tidyverse"
author: "Michael Robinson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this assignment, I will explore a synthetic healthcare dataset using the TidyVerse packages. The dataset, designed to mimic real-world healthcare data, is comprehensive, containing 10,000 rows and 15 columns. It includes a range of information typical in healthcare settings, such as patient names, ages, genders, medical conditions, admission and discharge dates, and more. thtih data was obtain from <https://www.kaggle.com/datasets/prasad22/healthcare-dataset>

# Load the Dataset and Libraries needed

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(DT)

patient_data <- read_csv("https://raw.githubusercontent.com/MRobinson112/tidy_verse/main/healthcare_dataset.csv")


datatable(patient_data, options = list(scrollX = TRUE))


```

# Filtering

Filtering is important in data analysis, it is used to narrow dataset to the relevant data. This increases the accuracy and relevance of the analysis by removing unnecessary data.

```{r}

# filter all patients over 50
patients_over_50 <- patient_data %>%
  filter(Age > 50)
         
# filter all patients over 50 that are male
male_patients_over_50 <- patient_data %>%
  filter(Gender == "Male", Age > 50)
```

# Arranging

Arranging data is used in data analysis to aid in understanding trends and patterns.

```{r}

# sort patients by date of admission 

patients_sorted_by_admission <- patient_data %>%
  arrange(`Date of Admission`)

head(patients_sorted_by_admission)
```

# Summarizing

Summarizing is important in data analysis because it helps to manage large datasets, find patterns, compare different data, and make decisions.

```{r}

average_billing_by_condition <- patient_data %>%
  group_by(`Medical Condition`) %>%
  summarize(Average_Billing = mean(`Billing Amount`))

average_billing_by_condition
```

# Visualization

visualization transforms data into a more accessible and interpretable form, enhancing understanding, communication, and decision-making.

```{r}

# Count of male patients over 50
count_male_over_50 <- nrow(male_patients_over_50)

# Count of all patients over 50
count_patients_over_50 <- nrow(patients_over_50)

# create a new data frame of summarized information.

plot_data <- data.frame(
  Category = c("Male Patients Over 50", "All Patients Over 50"),
  Count = c(count_male_over_50, count_patients_over_50)
)

ggplot(plot_data, aes(x = Category, y = Count, fill = Category)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Comparison of Male Patients Over 50 vs. All Patients Over 50",
       x = "",
       y = "Count")

```

# Extending the code

Your task here is to Extend an Existing Example. Using one of your classmate's examples (as created above), extend his or her example with additional annotated code.

## Answering additional questions

I thought this dataset was interesting and wanted to analyze some other questions I had. I will use mostly the same functions used in the original vignette to answer some other questions.

Healthcare cost can be extremely expensive. This is a huge problem for many people and is something that needs to be addressed. However, elective procedures are typically not strictly necessary and therefor their costs are less of a concern.

Here, I plotted the average costs of healthcare in the dataset by the different types of procedures.

```{r}
ggplot(patient_data, aes(x = `Admission Type`, y = `Billing Amount`, fill = `Admission Type`)) +
  geom_bar(stat = "summary", fun = "mean") +
  labs(title = "Bar Plot of Billing Price by Admission Type",
     x = "Admission Type",
     y = "Average Billing Price") +
  theme_minimal() +
  theme(legend.position = "none")
```

Another thing I wanted to see was whether the insurance provided has an impact on the price billed. I wanted to plot them divided by elective procedures and non-elective ones (both urgent and emergency). Using the same codeas in the vignette, I did that.

```{r}
# filter for elective procedures
elective_procedures <- patient_data %>%
  filter(`Admission Type` == "Elective")
         
# filter for non-elective procedures
urgent_procedures <- patient_data %>%
  filter(`Admission Type` == "Emergency" | `Admission Type` == "Urgent")
```

```{r}
#summarize the average billing amount for both subsets
average_billing_elective <- elective_procedures %>%
  group_by(`Insurance Provider`) %>%
  summarize(Average_Billing = mean(`Billing Amount`)) 

average_billing_urgent <- urgent_procedures %>%
  group_by(`Insurance Provider`) %>%
  summarize(Average_Billing = mean(`Billing Amount`))
```

```{r}
#combine the two subsets to help plot them
combined_average <- bind_rows(elective = average_billing_elective, urgent = average_billing_urgent, .id = "Procedure_Type")
```

```{r}
#plot the results
ggplot(combined_average, aes(x = `Insurance Provider`, y = Average_Billing, fill = Procedure_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Differences Between Urgent and Elective Procedure Prices",
       x = "Insurance Provider", y = "Average Billing Amount") +
  theme_minimal()
```

## Conclusion
There doesn't seem to be any major price differences between different insurances. Additionally, the average costs of elective procedures does seem to be higher than non-elective procedures accross all insurance providers. This makes sense to me.
